---
title: "Reproducible Research: Peer Assessment 1"
output:
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
### First we shall check if the file exists.
```{r LoadAndPreprocess, echo=TRUE}
if (!file.exists("activity/activity.csv")) {
        print("activity csv does not exist")
}
df <- read.csv("activity/activity.csv", stringsAsFactors = FALSE)
df$date <- as.Date(df$date, "%Y-%m-%d")
summary(df)
str(df)
head(df)
```


## What is mean total number of steps taken per day?
### We will perform the sum using the aggregate function
### Next will calculate the mean and median
```{r totalSteps}
library(dplyr)
library(ggplot2)
# Q 1.1
steps <- aggregate(df$steps,by = list(df$date), FUN = "sum")
colnames(steps) <- c("date","totalSteps")
steps
```
### The histogram below shows the frequency of total number of steps between 10000 and 15000 is more than 25 times in the two months
```{r totalStepsHistogram}
# Q 1.2
steps$row <- rownames(steps)
steps$row <- as.numeric(steps$row)
hist(steps$totalSteps, main = "Histogram of total steps", xlab = "Total Steps")

# Q 1.3
mean(steps$totalSteps, na.rm = TRUE);median(steps$totalSteps, na.rm = TRUE)

```


## What is the average daily activity pattern?
### The average daily pattern below shows that the activity seems to be high around the interval 835( 8:00 - 9:00 am) minutes
```{r averageActivityPattern}
stepsInterval <- group_by(df,interval) %>% summarise(avgSteps = mean(steps, na.rm = TRUE))
plot(stepsInterval$interval, stepsInterval$avgSteps, type="l", xlab= "Interval", ylab= "Avg steps for the interval for all days", main = "Average Steps per 5 minute interval for all days")
stepsInterval[which(stepsInterval$avgSteps==max(stepsInterval$avgSteps)), "interval"]

```



## Imputing missing values
### There are around 2304 values missing and shown as NAs
```{r imputeMissingValues}
length(which(is.na(df)))
dfNew <- df
naRows <- which(is.na(df[,1]))
length(naRows)
```

### Substituting the missing values with the average steps during that 5 minute interval
```{r}
for(i in 1 : length(naRows))
{
        k <- naRows[i]
        if(is.na(dfNew[k, 1]))
           {
        dfNew[k, "steps"] <- stepsInterval[which(stepsInterval[, "interval"]==dfNew[k, "interval"]), "avgSteps"]
        }
}
head(dfNew)
```

### Drawing a histogram with the imputed values shows that the frequency of the number of steps between 10000 and 15000 is 35
### The mean and median have also significantly increased
```{r totalStepsHistogramWithImputedValues}
stepsNew <- aggregate(dfNew$steps,by = list(dfNew$date), FUN = "sum")
colnames(stepsNew) <- c("date","totalSteps")
stepsNew
stepsNew$row <- rownames(stepsNew)
stepsNew$row <- as.numeric(stepsNew$row)
hist(stepsNew$totalSteps, main = "Histogram of total steps(by imputing missing data)", xlab = "Total steps", ylab = "Frequency")
mean(stepsNew$totalSteps, na.rm = TRUE);median(stepsNew$totalSteps, na.rm = TRUE)
```



## Are there differences in activity patterns between weekdays and weekends?
### Yes, the time series plot below shows the average steps in the 5 minute interval are higher in the weekend than on weekdays. The imputed data is used for these calculations
```{r timeSeriesPlot}
library(timeDate)
getIsWeekendString <- function(x)
{
    if(isWeekday(x, wday=1:5))
    {
            return("weekday");
    }else
    {
            return("weekend");
    }
}
dfNew$dow <- as.factor(sapply(dfNew$date, getIsWeekendString))

par(mfrow=c(2,1))

dfNewWkday <- dfNew[which(dfNew$dow=="weekday"), ]
dfNewWkend <- dfNew[which(dfNew$dow=="weekend"), ]

stepsIntWkday <- group_by(dfNewWkday,interval) %>% summarise(avgSteps = mean(steps, na.rm = TRUE))
stepsIntWkend <- group_by(dfNewWkend,interval) %>% summarise(avgSteps = mean(steps, na.rm = TRUE))

stepsIntWkday$dow <- "weekday"
stepsIntWkend$dow <- "weekend"
stepsInt <- rbind(stepsIntWkday, stepsIntWkend)
stepsInt$dow <- as.factor(stepsInt$dow)

library(lattice)
xyplot(avgSteps~interval| dow,
       data = stepsInt,
       type = "l",
       layout=c(1,2)
      )
```



